Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-06 paper.tex

Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded. In this section we describe the data transformation procedure and data visualisation technique employed in our case study.

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualisation.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.

@@

Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
Of these techniques, flow visualisations are commonly employed to facilitate the comparison of aggregated movement. 
Flow visualisations can be broadly categorized into two groups. Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
Since our data solely consists of OD information, we narrow our description to the latter group of visualisation techniques.

@@

The chord diagram is a visualisation technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
Chords or curved line segements are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualisation and details are provided in popup dialogue when upon mouse-over.

@@

The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
Visualising our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
The reverse occurs in the evening when trips are made towards provincial stations.
Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liege, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven. 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with   Brussels as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.

@@

We also see weird things happening in Antwerp, where a large amount of people go to the province in the morning on weekdays, while the oposite happens in Liege.
When looking into the data, the Antwerp pattern appears due to one specific route which is queried each morning.
Is there a group of colleagues all using the same app taking the same train daily?
Is it someone who checks this trip over and over again for delays?
We can only guess\ldots

>>>>

Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded. In this section we describe the data transformation procedure and data visualization technique employed in our case study.

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.

@@

Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
Flow visualizations can be broadly categorized into two groups. Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.

@@

The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
Chords or curved line segements are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualization and details are provided in popup dialogue when upon mouse-over.

@@

The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
The reverse occurs in the evening when trips are made towards provincial stations.
Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liege, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven. 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with   Brussels as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.

@@

There are several inconsistent patterns in the data that do not correspond with existing understanding of flows in Belgium.
In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
